\section{Formative review of artifacts}\label{FormativeStudy}

Head et al.'s prior content analysis~\cite{Head2022-ws} identified four patterns of interactive formula augmentation: scrubbable expressions, editable text fields, linked selections, and external controls. Head et al. were focused on augmentations broadly. Give our our focus on interaction, we sought additional documents with this pattern. We aimed to identify common forms of augmentation and better understand potential frictions in their implementation.

We acquired a set of 11 documents containing formulas of interest. Documents came from sources including Distill.pub, Setosa.io, Explorable Explanations, complex-analysis.com, Georgia Tech's Interactive Linear Algebra textbook, and other personal websites. Sources were curated from \citet{Head2022-ws}'s content analysis, \url{https://explorabl.es/}, \url{https://github.com/ubavic/awesome-interactive-math}, and our own web searches. Two authors found and analyzed 46 formulas of interest. All included formulas were either themselves dynamic or adjacent to and significant in some interactive mathematical explanation.

\paragraph{Augmentation patterns}
Of these formulas, 24 were dynamic, themselves changing following interaction. 22 supported some kind of linking between formula elements and visualization elements. Formulas were controlled by sliders (8), draggable points in graphs (17), and radio buttons (8), and through scrubbing (5) and editing (1) in the formula itself. Walkthroughs---step-by-step navigation through staged derivations---were used for formulas in 1 example, and additionally for mathematical content broadly in 2 others. We further reviewed the code involved in the open source repositories we were able to find for 11 documents. We noticed a number of recurring implementation patterns which might be potential frictions (see also appendix Table~\ref{tab:formative-audit}). These are:

\begin{itemize}
\item \emph{F1: Diffuseness and tangling (11 articles).} A single interaction is implemented across multiple files; or rendering, layout, and behavior is tangled within a single function.
\item \emph{F2: Indirect identifiers (11 articles).} Variables are referred to in the code using identifiers or array indexing distinct from how the variable appears in the formula.
\item \emph{F3: Fussing the DOM and LaTeX (9 articles)}: The code constructed \LaTeX{} strings programmatically, traversed rendered formula DOM, or both. The two exceptions bypassed LaTeX with their own rendering engines.
\item\emph{F4: Repetition (10 articles)}: interaction logic appeared duplicated across instances of related math objects.
\item\emph{F5: State management (10 articles)}: Articles implemented custom state management and wired up drag handlers, click listeners, and other callbacks to propagate changes across formulas, variables, and other representations.
\end{itemize}

We posit these potential frictions could place distance between an author and focusing on the work of presenting a formula well. Our later controlled study (Section~\ref{results}) compares Delta to a baseline to see if improved tooling can reduce the work in these task.